Conversations in time: interactive visualisation to explore structured temporal data

### 2020-08-24 tsibbletalk.Rmd
introducing plotly_key_tree()

```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = 'figure/', fig.align = 'center', fig.show = 'hold',
  cache = TRUE, cache.path = 'cache/',
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
opts_knit$set(root.dir = here::here())
read_chunk("scripts/demo.R")
```

@@

To incorporate structured specifications in the `key`, a symbolic formula can be passed to the `spec` argument. Adopting Wilkinson notations for factorial models [@Wilkinson1973], the `spec` follows the `/` and `*` operators tradition to declare nesting and crossing variables respectively. The `spec` for the `aus_retail` data is therefore specified as `spec = State * Industry`, which is the default for the presence of multiple `key` variables. Using `/` indicates if there is a hierarchy in the data. `tourism_monthly`

Specifying the structure is particularly useful, when there's a hierarchical structure in the data. For such case, a new function `plotly_key_tree()` can help to visualise the "key" structure as a dendrogram, using plotly. This type of plot gives an overview of structured information in the data. Owing to its interactivity, each node in the tree plot is clickable and linked to other plots, and in turn the navigation at various levels is made easier.

* Constructing recursive lists. Lists of key values, and json
* Self-linking: a single data source

>>>>

```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE, comment = "#>",
  fig.path = 'figure/', fig.align = 'center', fig.show = 'hold',
  cache = TRUE, cache.path = 'cache/',
  out.width = ifelse(is_html_output(), "100%", "\\textwidth")
)
opts_knit$set(root.dir = here::here())
read_chunk("scripts/demo.R")
read_chunk("scripts/tourism.R")
```

@@

To incorporate structured specifications in the `key`, a symbolic formula can be passed to the `spec` argument. Adopting Wilkinson notations for factorial models [@Wilkinson1973], the `spec` follows the `/` and `*` operators tradition to declare nesting and crossing variables respectively. The `spec` for the `aus_retail` data is therefore specified as `State * Industry` or `Industry * State`, which is the default for the presence of multiple `key` variables. If there is a hierarchy in the data, using `/` is required to indicate the parent-child relation, as strictly one direction `parent/child`.

The `tourism_monthly` dataset packaged in \CRANpkg{tsibbletalk}, contains monthly domestic overnight trips across Australia, to give an illustrator of nesting and crossing. The `key` is comprised of three identifying variables: `State`, `Region`, and `Purpose` (of trip), in particular `State` nesting of `Region`, together crossed with `Purpose`. This specification can be translated as follows:

```{r tourism-shared, echo = TRUE}
```

```{r load-other-pkgs}
```

This dataset contains a three-level hierarchy: the root node is implicitly Australia, and geographically disaggregated to states and lower-level tourism regions. A new function `plotly_key_tree()` has been implemented to address the need of hierarchical discovery arising from data. It visually interprets hierarchies in the shared tsibble's `spec` as a tree view, made with \CRANpkg{plotly}. From a static point of view, a tree visualisation of hierarchical data facilitates to grasp the overview structure. From the interactive perspective, selecting a parent node along the tree subsequently triggers its children and their corresponding series, or vice versa. It makes navigation simpler than conditional panels, and also enables persistent brushing for comparisons across groups.

```{r tourism-linking}
include_graphics("img/tourism-linking.png")
```

```{r plotly-key-tree, eval = FALSE}
```

### 2020-08-24 tsibbletalk.tex
introducing plotly_key_tree()

In the spirit of tidy data to the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, in order to build a
fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} and \CRANpkg{fable} packages
fill the role of statistical analysis and forecasting in the
\textbf{tidyverts} ecosystem. When time series analysis starts taking
off, series of interest denoted by the \texttt{key} variables often
remain unchanged over the course of analysis, from trend inspection to
forecasting performance. Figure \ref{fig:highlight-retail} (a) gives an
overview of 152 series for the retail data using an overlaid time series
plot, while Figure \ref{fig:highlight-retail} (b) presents a
scatterplot, where each series is represented by a dot in the feature
space (trend versus seasonal strength). The plot making of Figure
\ref{fig:highlight-retail} (b) is aided with the \texttt{features()}
function from \CRANpkg{feasts}, which summarises original data by each
series down to various statistical features. This function along with
other \textbf{tidyverts} functions is \texttt{tsibble}-aware, and
outputs a table in a reduced form where each row corresponds to a
series, thus graphically displayed as Figure \ref{fig:highlight-retail}
(b).

@@

To incorporate structured specifications in the \texttt{key}, a symbolic
formula can be passed to the \texttt{spec} argument. Adopting Wilkinson
notations for factorial models \citep{Wilkinson1973}, the \texttt{spec}
follows the \texttt{/} and \texttt{*} operators tradition to declare
nesting and crossing variables respectively. The \texttt{spec} for the
\texttt{aus\_retail} data is therefore specified as
\texttt{spec\ =\ State\ *\ Industry}, which is the default for the
presence of multiple \texttt{key} variables. Using \texttt{/} indicates
if there is a hierarchy in the data. \texttt{tourism\_monthly}

Specifying the structure is particularly useful, when there's a
hierarchical structure in the data. For such case, a new function
\texttt{plotly\_key\_tree()} can help to visualise the ``key'' structure
as a dendrogram, using plotly. This type of plot gives an overview of
structured information in the data. Owing to its interactivity, each
node in the tree plot is clickable and linked to other plots, and in
turn the navigation at various levels is made easier.

\begin{itemize}
\tightlist
\item
  Constructing recursive lists. Lists of key values, and json
\item
  Self-linking: a single data source
\end{itemize}

@@

\address{%
Earo Wang\\
The University of Auckland\\
Department of Statistics\\ The Univerisity of Auckland\\ New Zealand\\
}
\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}

>>>>

In the spirit of tidy data to the \CRANpkg{tidyverse}
\citep{Wickham2019}, the \textbf{tidyverts} suite features
\texttt{tsibble} as the foundational data structure, in order to build a
fluid and fluent pipeline for time series analysis. Besides
\CRANpkg{tsibble}, the \CRANpkg{feasts} and \CRANpkg{fable} packages
fill the role of statistical analysis and forecasting in the
\textbf{tidyverts} ecosystem. When time series analysis starts taking
off, series of interest denoted by the \texttt{key} variables often
remain unchanged over the course of analysis, from trend inspection to
forecasting performance.

Figure \ref{fig:highlight-retail} (a) gives an overview of 152 series
for the retail data using an overlaid time series plot, while Figure
\ref{fig:highlight-retail} (b) presents a scatterplot, where each series
is represented by a dot in the feature space (trend versus seasonal
strength). The plot making of Figure \ref{fig:highlight-retail} (b) is
aided with the \texttt{features()} function from \CRANpkg{feasts}, which
summarises original data by each series down to various statistical
features. This function along with other \textbf{tidyverts} functions is
\texttt{tsibble}-aware, and outputs a table in a reduced form where each
row corresponds to a series, thus graphically displayed as Figure
\ref{fig:highlight-retail} (b).

@@

To incorporate structured specifications in the \texttt{key}, a symbolic
formula can be passed to the \texttt{spec} argument. Adopting Wilkinson
notations for factorial models \citep{Wilkinson1973}, the \texttt{spec}
follows the \texttt{/} and \texttt{*} operators tradition to declare
nesting and crossing variables respectively. The \texttt{spec} for the
\texttt{aus\_retail} data is therefore specified as
\texttt{State\ *\ Industry} or \texttt{Industry\ *\ State}, which is the
default for the presence of multiple \texttt{key} variables. If there is
a hierarchy in the data, using \texttt{/} is required to indicate the
parent-child relation, as strictly one direction \texttt{parent/child}.

The \texttt{tourism\_monthly} dataset packaged in \CRANpkg{tsibbletalk},
contains monthly domestic overnight trips across Australia, to give an
illustrator of nesting and crossing. The \texttt{key} is comprised of
three identifying variables: \texttt{State}, \texttt{Region}, and
\texttt{Purpose} (of trip), in particular \texttt{State} nesting of
\texttt{Region}, together crossed with \texttt{Purpose}. This
specification can be translated as follows:

\begin{Schunk}
\begin{Sinput}
library(tsibbletalk)
tourism_shared <- tourism_monthly %>% 
  as_shared_tsibble(spec = (State / Region) * Purpose)
\end{Sinput}
\end{Schunk}

This dataset contains a three-level hierarchy: the root node is
implicitly Australia, and geographically disaggregated to states and
lower-level tourism regions. A new function \texttt{plotly\_key\_tree()}
has been implemented to address the need of hierarchical discovery
arising from data. It visually interprets hierarchies in the shared
tsibble's \texttt{spec} as a tree view, made with \CRANpkg{plotly}. From
a static point of view, a tree visualisation of hierarchical data
facilitates to grasp the overview structure. From the interactive
perspective, selecting a parent node along the tree subsequently
triggers its children and their corresponding series, or vice versa. It
makes navigation simpler than conditional panels, and also enables
persistent brushing for comparisons across groups.

\begin{Schunk}

\begin{center}\includegraphics[width=\textwidth]{img/tourism-linking} \end{center}

\end{Schunk}

@@

\address{%
Earo Wang\\
The University of Auckland\\
Department of Statistics\\ The University of Auckland\\ New Zealand\\
}
\href{mailto:earo.wang@auckland.ac.nz}{\nolinkurl{earo.wang@auckland.ac.nz}}
